prolfqua - A Comprehensive R-package for Proteomics Differential Expression Analysis

### 2022-06-07 inst/MyArticle/paper.Rmd
replace condition with group

---
title: prolfqua: A Comprehensive R-package for Proteomics Differential Expression Analysis
author: 
  - Witold E. Wolski $^{\dagger\ddagger}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - [University of Zurich](https://uzh.ch)/[ETH Zurich](https://ethz.ch), Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. $^\ddagger$[Swiss Institute of Bioinformatics (SIB), Quartier Sorge - Batiment Amphipole, 1015 Lausanne, Switzerland](https://www.sib.swiss/) $^\dagger$Correspondence \texttt{wew@fgcz.ethz.ch}]
  - Paolo Nanni$^{\ast}$
  - Jonas Grossmann$^{\ast\ddagger}$
  - Maria d'Errico$^{\ast\ddagger}$
  - Ralph Schlapbach$^{\ast}$
  - Christian Panse$^{\ast\ddagger}$
output: 
  bookdown::pdf_document2:
      toc: true
      toc_appendix: false
      toc_bib: false
      keep_tex: true
date: "`r format(Sys.time())`"
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
abstract: |
    Mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. Nevertheless, there is a need for a flexible and easy-to-use application programming interface in R that transparently supports a variety of well principled statistical procedures. The `r BiocStyle::Githubpkg("fgcz/prolfqua")` package can model simple experimental designs with a single explanatory variable and complex experiments with multiple factors and hypothesis testing. It integrates essential steps of the mass spectrometry-based differential expression analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. The application programmer interface strives to be clear,  predictable, discoverable, and consistent to make proteomics data analysis easy and exciting. 
     Furthermore, the package implements benchmark functionality that can help to compare data acquisition, data preprocessing, or data modeling methods using a gold standard dataset. Finally, we show that the implemented methods allow sensitive and specific differential expression analysis. 
     The `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package is available on GitHub https://github.com/fgcz/prolfqua, distributed under the MIT licence, and runs on all platforms supported by the [R free software environment for statistical computing and graphics](https://www.r-project.org/).
urlcolor: blue
---

@@

An important aspect of mass spectrometric data are missing peptide and protein quantifications. @rubin1976inference classified missing data problems into three categories: missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR). For instance, in data-dependent acquisition (DDA) MS, only a limited number of MS1 signals are selected for fragmentation and identified. Peptide quantification algorithms transfer identification information between MS1 features in different samples by aligning the data using retention time and mass information, thus reducing the amount of missing data. Nevertheless, highly abundant proteins can suppress the detection of other proteins in a sample, a MAR mechanism. Furthermore, a weak correlation between the number of missing measurements in a condition and the abundance of the quantified protein can be observed, which is caused by the limit of detection (LOD), a MNAR mechanism [@mcgurk2020use].

@@

When developing the R package `r BiocStyle::Githubpkg("fgcz/prolfqua")` we were inspired by the R package `r BiocStyle::CRANpkg("caret")` [@caret2008], which enables us to call various machine learning (ML) methods, and makes selecting the best ML algorithm for your problem easy. We aimed for a similar R package for the differential expression analysis of proteomics data. However, after examining the R packages for modeling proteomics quantification data, we observed that model specification, input, and output formats differ widely. At the same time, they have in common that: they fit linear models either to peptide or protein intensities, determine contrasts among conditions, and afterward apply empirical Bayes variance shrinkage. Therefore, the revised goal was to provide a modular object-oriented (OO) design, with R linear models as a core, where we can add features such as $p$-value aggregation, e.g., `r BiocStyle::Biocpkg("PECA")`, variance shrinkage, or modeling of missing observations.

@@

```{r exampleContrasts, fig.cap="(ref:exampleContrasts)", echo=TRUE, out.width = '90%', fig.width = 8, fig.height = 5}
# specify differences among conditions
contrasts <- c(
  "dilution_(9/7.5)_1.2" =   "dilution.e - dilution.d",
  "dilution_(7.5/6)_1.25" =   "dilution.d - dilution.c"
)
## fit model
lmmodel <- paste(lfqp$intensity_column()," ~ dilution.")
modelFunction <- prolfqua::strategy_lm( lmmodel, model_name = "lm")
models <- prolfqua::build_model(lfqp, modelFunction)

@@

(ref:FDRfdp) Panel A - Number of estimated contrasts for each modeling method (higher is better). Panel B - Partial area under the ROC curve at $10\%$ FPR ($pAUC_10$) for all contrasts and three different statistics: the difference among conditions, the scaled $p$-value (sign(diff) $\cdot$ p.value) and the $t$-statistics (higher is better). The red line indicates the average area under the curve of all methods.  Panel C - Plots the false discovery proportion (FDP) as a function of the FDR. Ideally, the FDR should be equal to the FDP. Therefore larger distances from the diagonal are worse.

>>>>

---
title: prolfqua - A Comprehensive R-package for Proteomics Differential Expression Analysis
author: 
  - Witold E. Wolski $^{\dagger\ddagger}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - [University of Zurich](https://uzh.ch)/[ETH Zurich](https://ethz.ch), Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. $^\ddagger$[Swiss Institute of Bioinformatics (SIB), Quartier Sorge - Batiment Amphipole, 1015 Lausanne, Switzerland](https://www.sib.swiss/) $^\dagger$Correspondence \texttt{wew@fgcz.ethz.ch}]
  - Paolo Nanni$^{\ast}$
  - Jonas Grossmann$^{\ast\ddagger}$
  - Maria d'Errico$^{\ast\ddagger}$
  - Ralph Schlapbach$^{\ast}$
  - Christian Panse$^{\ast\ddagger}$
output: 
  bookdown::pdf_document2:
      toc: true
      toc_appendix: false
      toc_bib: false
      keep_tex: true
date: "`r format(Sys.time())`"
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
abstract: |
    Mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. Nevertheless, there is a need for a flexible and easy-to-use application programming interface in R that transparently supports a variety of well principled statistical procedures. The `r BiocStyle::Githubpkg("fgcz/prolfqua")` package can model simple experimental designs with a single explanatory variable and complex experiments with multiple factors and hypothesis testing. It integrates essential steps of the mass spectrometry-based differential expression analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. The application programmer interface strives to be clear,  predictable, discoverable, and consistent to make proteomics data analysis easy and exciting. 
     Furthermore, the package implements benchmark functionality that can help to compare data acquisition, data preprocessing, or data modeling methods using a gold standard dataset. Finally, we show that the implemented methods allow sensitive and specific differential expression analysis. 
     The `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package is available on GitHub https://github.com/fgcz/prolfqua, distributed under the MIT licence, and runs on all platforms supported by the [R free software environment for statistical computing and graphics](https://www.r-project.org/).
urlcolor: blue
---

@@

An important aspect of mass spectrometric data are missing peptide and protein quantifications. @rubin1976inference classified missing data problems into three categories: missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR). For instance, in data-dependent acquisition (DDA) MS, only a limited number of MS1 signals are selected for fragmentation and identified. Peptide quantification algorithms transfer identification information between MS1 features in different samples by aligning the data using retention time and mass information, thus reducing the amount of missing data. Nevertheless, highly abundant proteins can suppress the detection of other proteins in a sample, a MAR mechanism. Furthermore, a weak correlation between the number of missing measurements in a group and the abundance of the quantified protein can be observed, which is caused by the limit of detection (LOD), a MNAR mechanism [@mcgurk2020use].

@@

When developing the R package `r BiocStyle::Githubpkg("fgcz/prolfqua")` we were inspired by the R package `r BiocStyle::CRANpkg("caret")` [@caret2008], which enables us to call various machine learning (ML) methods, and makes selecting the best ML algorithm for your problem easy. We aimed for a similar R package for the differential expression analysis of proteomics data. However, after examining the R packages for modeling proteomics quantification data, we observed that model specification, input, and output formats differ widely. At the same time, they have in common that: they fit linear models either to peptide or protein intensities, determine differences among groups, and afterward apply empirical Bayes variance shrinkage. Therefore, the revised goal was to provide a modular object-oriented (OO) design, with R linear models as a core, where we can add features such as $p$-value aggregation, e.g., `r BiocStyle::Biocpkg("PECA")`, variance shrinkage, or modeling of missing observations.

@@

```{r exampleContrasts, fig.cap="(ref:exampleContrasts)", echo=TRUE, out.width = '90%', fig.width = 8, fig.height = 5}
# specify differences among groups
contrasts <- c(
  "dilution_(9/7.5)_1.2" =   "dilution.e - dilution.d",
  "dilution_(7.5/6)_1.25" =   "dilution.d - dilution.c"
)
## fit model
lmmodel <- paste(lfqp$intensity_column()," ~ dilution.")
modelFunction <- prolfqua::strategy_lm( lmmodel, model_name = "lm")
models <- prolfqua::build_model(lfqp, modelFunction)

@@

(ref:FDRfdp) Panel A - Number of estimated contrasts for each modeling method (higher is better). Panel B - Partial area under the ROC curve at $10\%$ FPR ($pAUC_10$) for all contrasts and three different statistics: the difference among groups, the scaled $p$-value (sign(diff) $\cdot$ p.value) and the $t$-statistics (higher is better). The red line indicates the average area under the curve of all methods.  Panel C - Plots the false discovery proportion (FDP) as a function of the FDR. Ideally, the FDR should be equal to the FDP. Therefore larger distances from the diagonal are worse.
